---
title: "markdown_practice"
output: pdf_document
date: "2023-01-15"
---

# Load libraries
### Load in tidyverse and janitor libraries
```{r, echo=FALSE}
# Installing a library
# install.packages("tidyverse")
# install.packages("janitor")

# Loading a library
library(tidyverse)
library(janitor)
```

# Read csv file
### Load a csv into a dataframe [Titanic dataset]
```{r}
titanic_file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/Discussion/R_Basics/titanic.csv'
# read.csv method
titanic = read.csv(titanic_file_path, na.strings='')

# view the dataframe
glimpse(titanic)
```
# Data Visualization
## Bar Chart
```{r}
# Bargraph of each of the Embarked classes
ggplot(
  data=titanic,
  aes(x=as.factor(Embarked)) # aes is short for aesthetics
) +
geom_bar(na.rm=TRUE)
```
\newpage

## Historgram
```{r}
# Histogram of the fare paid 
ggplot(
  data=titanic,
  aes(x=Fare)
) +
geom_histogram()
```
\newpage

## Bargraph
```{r}
# Bargraph of each of the Embarked classes with survived amount
ggplot(
  data=titanic,
  aes(x=as.factor(Embarked),
      fill=as.factor(Survived))
) + geom_bar()
```

\newpage
# Dataframe data transformations [Penguins dataset]
```{r}
penguins_file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/Discussion/R_Basics/penguins_full.csv'
penguins = read.csv(penguins_file_path, na.strings='')
glimpse(penguins)
```

## chr -> numeric data conversion
**bill_length_mm** and **bill_depth_mm** need to be converted to numeric data types
```{r}
penguins$bill_length_mm=as.numeric(penguins$bill_length_mm)
penguins$bill_depth_mm = as.numeric(penguins$bill_depth_mm)
glimpse(penguins)
```
\newpage
## chr (categorical) -> factor (numerical representation of class value)
**species** a class variable needs to be converted to a factor variable
```{r}
penguins$species=as.factor(penguins$species)
glimpse(penguins)
```
\newpage
# Data Visualizations with transformed penguins dataframe
## Boxplot
```{r}
# Boxplot of bill lengths for each species type (boxplot plots the 5 number summary)
ggplot(
  data=penguins,
  aes(x=species,
      y=bill_length_mm)
) +
geom_boxplot()
```

\newpage
## Scatterplot
```{r}
# Scatterplot of bill depth and bill length, marked by each species
ggplot(
  data=penguins,
  aes(x=bill_depth_mm,
      y=bill_length_mm,
      shape=species,
      color=species)
) +
geom_point()

# Making mark of each species larger in previous plot
ggplot(
  data=penguins,
  aes(x=bill_depth_mm,
      y=bill_length_mm,
      shape=species,
      color=species)
) +
geom_point(size=4)
```

\newpage
# The tilde (~) and tilde dot (~. or .~) operators

The variable on the **left-hand side of tilde operator is the dependent variable** and the variable(s) on the **right-hand side of tilde operator is/are called the independent variable(s)**. 

For example:
Regression_Model <- lm(y~ x1 + x2 + x3)

If want to include all of a dataframe's columns, then can combine the tild operator with a dot (.). 

For example:
Regression_Model_New < - lm(y~ . , data = Regression_Data)

## Stacking scatterplots horizontally
```{r}
# Stack the scatterplots horizontally  
ggplot(
  data=penguins,
  aes(x=bill_depth_mm,
      y=bill_length_mm,
      shape=species,
      color=species)
) +
geom_point(size=4) + facet_grid(.~species)
```
\newpage
## Stacking scatterplots vertically
```{r}
ggplot(
  data=penguins,
  aes(x=bill_depth_mm,
      y=bill_length_mm,
      shape=species,
      color=species)
) +
geom_point(size=4) + facet_grid(species~.)
```

\newpage
# Dataframe methods & Chaining methods [LAPD dataset]

## loading in dataset
```{r}
lapd_file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/Discussion/R_Basics/Police_Payroll.csv'
lapd = read.csv(lapd_file_path, na.strings='')

glimpse(lapd)
```

\newpage
## Cleaning dataframe column names
```{r}
lapd = clean_names(lapd)
glimpse(lapd)
```

\newpage
## head() method
```{r}
# We do head() to view only the first few lines of the data
head(select(lapd, year, base_pay))
```

## Selecting specific columns of a dataframe
```{r, results="hide"}
lapd %>%
 select(year, base_pay)
```

## Removing specific columns of a dataframe 
```{r}
# selects all of the columns except the ones with negative sign inside select() method
head(select(lapd, -row_id, -department_title))
```

\newpage
## starts_with() method
```{r}
head(select(lapd, starts_with('q')))
```

## ends_with() method
```{r}
head(select(lapd, ends_with('pay')))
```

\newpage
## Slicing - retrieving specific rows
```{r}
slice(lapd, 3:7)
```

## Filerting - filer columns based on boolean predicate
```{r, results="hide"}
filter(lapd, year == 2018)
```

\newpage
## Method chaining
```{r, results="hide"}
# filter lapd on year = 2018 and base_pay > 62474
lapd %>%
  filter(year == 2018 & base_pay > 62474)

# filter lapd on year = 2018 and base_pay > 62474 and display the row counts
lapd %>%
  filter(year == 2018 & base_pay > 62474) %>%
  nrow()

# filter lapd on year >= 2013 and year <= 2015 and display the row counts
lapd %>%
  filter(year >= 2013 & year <= 2015) %>%
  nrow()

# filter lapd on employment_type = full time and year = 2018 and display the row counts
lapd %>%
  filter(employment_type == "Full Time" & year == 2018) %>%
  nrow()
```

### Method chaining - 2 chains
```{r, results="hide"}
# apply a filter to lapd and then select specific columns of the filtered dataframe
lapd %>%
  filter(year == 2018) %>%
  select(job_class_title,
         employment_type,
         base_pay)
```

## Dataframe transformation of data with mutate() method
```{r, results="hide"}
# divide every base_pay_k value by 1000
lapd %>%
  mutate(base_pay_k = base_pay/1000)
```

### View summary of mutated lapd dataframe
```{r}
summarize(lapd, mean_base_pay = mean(base_pay)) 
```

## Creating a new column with boolean logic
```{r, results="hide"}
# Create a new variable called base_pay_level which has Less Than 0, No Income, Less than Median and Greater than 0 and Greater than Median. We will consider $62474 as the median
lapd %>% 
  mutate(base_pay_level = case_when(
    base_pay < 0 ~ "Less than 0", 
    base_pay == 0 ~ "No Income",
    base_pay < 62474 & base_pay > 0 ~ "Less than Median, Greater than 0",
    base_pay > 62474 ~ "Greater than Median"))
```

## Aggregate dataframe data using group_by() method
```{r}
lapd %>% 
  group_by(employment_type) %>% 
  summarize(med_base_pay = median(base_pay),
            count = n())
```