---
title: "Depression Modeling"
author: "Tarek El-Hajjaoui"
date: "2023-02-25"
output: pdf_document
---

---
title: "MCAT Modeling & Analysis"
author: "Tarek El-Hajjaoui"
date: "2023-02-25"
output: pdf_document
---

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Loading libraries
library(tidyverse)
library(janitor)
library(ggplot2)
```

### Data loading & pre-processing

Loading the dataset
```{r}
file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/R_Statistical_Modeling/Depression/depression.txt'
df = read.table(file_path, header=TRUE, sep="", dec=".")
```

Summary of dataset - Note: Y = effectiveness score
```{r}
str(df)
```

Transform categorical columns to as.factor data types
```{r}
categorical_cols <- c('x2', 'x3', 'TRT')
df[categorical_cols] <- lapply(df[categorical_cols], as.factor)
```

Ensuring column data types are correct now.
```{r}
str(df)
```

\newpage

Model 1: Y =y (the effectiveness score) and X1=age
```{r}
model <- lm(y ~ age, data=df)
```

Model 1 summary
```{r}
summary(model)
```

\newpage

Model 2: Y=effectiveness score, X1=age, X2=TRT (treatment), X3=age*TRT
```{r}
model_2 <- lm(y ~ age + TRT + age*TRT, data=df)
```

Model 2 summary
```{r}
summary(model_2)
```
