---
title: "Heart Pulse"
author: "Tarek El-Hajjaoui"
date: "2023-02-09"
output: pdf_document
---

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Loading libraries
library(tidyverse)
library(janitor)
library(ggplot2)
```

Loading the dataset
```{r}
file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/R_Statistical_Modeling/HeartPulse/pulse.txt'
df = read.table(file_path, header=TRUE, sep="", dec=".")
```

Summary of data set
```{r}
str(df)
```

Correcting categorical column data types.
```{r}
df$Smoke = as.factor(df$Smoke)
df$Gender = as.factor(df$Gender)
df$Exercise = as.factor(df$Exercise)
```

Verifying column transformations
```{r}
str(df)
```

\newpage

Summary of dataset
```{r}
summary(df)
```

*Model Creation below

\newpage
Creating a model in which the response, Y, is Rest (resting heart rate), and the covariates, X, are Hgt (Height in inches), Wgt (Weight in pounds) and Smoke (smoking status - 1 for smokers and 0 for non-smokers), and an interaction between height and weight.

$$ \hat{Y} = \hat{Rest} $$
$$ X_1 = Hgt $$
$$ X_2 = Wgt $$
$$ X_3 = Smoke $$
$$ X_4 = Hgt * Wgt $$

## Multicorviate Linear Regression Full Model

$$\hat{Y} = \hat{\beta_0} + \hat{\beta_1}X_1 + \hat{\beta_2}X_2 + \hat{\beta_3}X_3 + \hat{\beta_4}(X_1*X_2)$$


```{r}
full_model <- lm(Rest ~ Hgt + Wgt + Smoke + Hgt * Wgt, data=df)
```

Summary of full model
```{r}
summary(full_model)
```

Calculating the SSE (Sum of Square Errors) for full model
```{r}
SSE <- sum(full_model$resid^2)
SSE
```

Comparing to built-in deviance( ) method
```{r}
deviance(full_model)
```

\newpage

## Multicorviate Linear Regression Reduced Model

$$\hat{Y} = \hat{\beta_0} + \hat{\beta_1}X_1 + + \hat{\beta_3}X_3 $$

```{r}
reduced_model = lm(Rest ~ Hgt + Smoke, data=df)
```

Using ANOVA to test difference between the reduced model and full model.
```{r}
anova(reduced_model, full_model)
```

Sequential Sum of Squares Regression on the full model via ANOVA function.

### Full Model

$$\hat{Y} = \hat{\beta_0} + \hat{\beta_1}X_1 + \hat{\beta_2}X_2 + \hat{\beta_3}X_3 + \hat{\beta_4}(X_1*X_2)$$
The Sequential Sum of Squares Regression on the full model is as follows:

$$ anova(fullmodel) \rightarrow SSR(X_1), SSR(X_1 | X_2), SSR(X_3 | X_1, X_2), SSR(X_4 | X_1, X_2, X_3) $$

```{r}
anova_table <- anova(full_model)
anova_table
```

Calculating the SSTO of the Sequential Sum of Squares Regression

$$ SSTO = SSR(Hgt) + SSR (Wgt | Hgt) + SSR(Smoke | Hgt, Wgt) + SSR(Hgt * Wgt | Hgt, Wgt, Smoke) + SSE $$

```{r}
SSTO <- sum(anova_table[, 2])
SSTO
```