---
title: "Airbnb NYC"
author: "Tarek El-Hajjaoui"
date: "2023-04-23"
output: html_document
---

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# library
library(tidyverse)
library(ggplot2)
library(dplyr)
```

### Q 4.11.3.1 Airbnb in NYC
Perform an EDA, build a model, and interpret model coefficients to describe variation in the number of reviews (a proxy for the number of rentals, which is not available) as a function of the variables provided. Don’t forget to consider an offset, if needed.

```{r}
# Loading the dataset
file_path = './NYCairbnb.csv'
df = read.csv(file_path, header = TRUE)
```

Explanatory Data Analysis (EDA)

```{r}
str(df)
```

Summary statistics
```{r}
# Some features of interest (except, number of reviews, is y)
multiple.func <- function(x) {
      round(c(summary = summary(x), var = var(x)), 2)
}
features <- c('number_of_reviews', 'price', 'days')
sapply(df[features], multiple.func)

features <- c('bedrooms', 'bathrooms')
sapply(df[features], multiple.func)

features <- c('review_scores_cleanliness', 'review_scores_location', 'review_scores_value')
sapply(df[features], multiple.func)
```

Preprocessing - transforming data types
```{r}
# Transforming categorical columns to factor data type.
categorical_cols <- c('room_type', 'bathrooms', 'bedrooms', 'review_scores_cleanliness', 'review_scores_location', 'review_scores_value', 'instant_bookable')
df[categorical_cols] <- lapply(df[categorical_cols], as.factor)

# Transforming date columns
date_cols <- c('last_scraped', 'host_since')
df[date_cols] <- lapply(df[date_cols], as.Date)
```

```{r}
# Ensure datatypes are correct now
str(df)
```

Checking if the data justifies a Poisson regression model
```{r}
# Create a matrix for plots
par(mfrow = c(1, 2))
plot(df[['price']], df$number_of_reviews, xlab='price', ylab="number_of_reviews", type = "h", xlim=c(0,2000))
plot(df[['days']], df$number_of_reviews, xlab='days', ylab="number_of_reviews", type = "h", xlim=c(0,3500))

par(mfrow = c(2, 3))
plot(df[['bedrooms']], df$number_of_reviews, xlab='bedrooms', ylab="number_of_reviews", type = "h")
plot(df[['bathrooms']], df$number_of_reviews, xlab='bathrooms', ylab="number_of_reviews", type = "h")
plot(df[['review_scores_cleanliness']], df$number_of_reviews, xlab='review_scores_cleanliness', ylab="number_of_reviews", type = "h")
plot(df[['review_scores_location']], df$number_of_reviews, xlab='review_scores_location', ylab="number_of_reviews", type = "h")
plot(df[['review_scores_value']], df$number_of_reviews, xlab='review_scores_value', ylab="number_of_reviews", type = "h")
```

Most of the graphs above suggest that the number of number of reviews follow a Poisson distribution (discussed more below).
