---
title: "Blood Pressure"
author: "Tarek El-Hajjaoui"
date: "2023-03-05"
output: pdf_document
---

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Loading libraries
library(tidyverse)
library(janitor)
library(ggplot2)
```

### Data loading & pre-processing

Loading the dataset
```{r}
file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/R_Statistical_Modeling/BloodPressure/Blood.txt'
df = read.table(file_path, header=TRUE, sep="", dec=".")
```

Summary of dataset
```{r}
str(df)
```

Transforming categorical columns to factor data type.
```{r}
categorical_cols <- c('Smoke', 'Overwt')
df[categorical_cols] <- lapply(df[categorical_cols], as.factor)
```

Ensuring column data types are correct now.
```{r}
str(df)
```

\newpage

Side by side boxplots comparing the blood pressure for the three weight groups.
```{r}
ggplot(df, aes(x=Overwt, y=SystolicBP, fill=Overwt)) + 
  geom_boxplot() + 
  xlab('Weight Class') +
  labs(caption = '0 = normal, 1 = overweight, 2 = obese')
```


The sample means, standard deviations and sample sizes for the three weight groups.

Weight Group 0
```{r}
# class 0: mean, std, sample size
df_0 <- subset(df, Overwt == 0)
str(df_0)
```

```{r}
mean_0 <- mean(df_0$SystolicBP)
sd_0 <- sd(df_0$SystolicBP)
n_0 <- length(df_0$SystolicBP)
stats_0 <- c(mean_0, sd_0, n_0)
print(stats_0)
```

\newpage

Weight Group 1
```{r}
# class 0: mean, std, sample size
df_1 <- subset(df, Overwt == 1)
str(df_1)
```

```{r}
mean_1 <- mean(df_1$SystolicBP)
sd_1 <- sd(df_1$SystolicBP)
n_1 <- length(df_1$SystolicBP)
stats_1 <- c(mean_1, sd_1, n_1)
print(stats_1)
```

Weight Group 2
```{r}
# class 0: mean, std, sample size
df_2 <- subset(df, Overwt == 2)
str(df_2)
```

```{r}
mean_2 <- mean(df_2$SystolicBP)
sd_2 <- sd(df_2$SystolicBP)
n_2 <- length(df_2$SystolicBP)
stats_2 <- c(mean_2, sd_2, n_2)
print(stats_2)
```

Conducting a one-way anova test

- One-way analysis of variance (ANOVA) is a statistical method for testing for differences in the means of three or more groups

Hypothesis Testing

- H0: a_0 == a_1 == a_2

- Ha: at least 1 mean difference, a_k, does not equal 0

```{r}
one_way_anova <- aov(SystolicBP ~ as.factor(Overwt), data=df)
summary(one_way_anova)
```
