---
title: "Understanding Simple Linear Regression Metrics"
output: pdf_document
date: "2023-01-26"
---

```{r, echo=FALSE, results="hide"}
# Loading libraries
library(tidyverse)
library(janitor)
```
Suppose a simple linear model is fit to predict Y= weight in kilograms using X= height in centimeters of an adult. But say a new simple linear model is fit using Y = height in centimeters and X= weight in kilograms (that is to say Y and X have reversed).
State whether each of the following would be the same for this new model as it was for the original model, or it would be different and explain in a sentence or two.

### Generate dummy data for testing
```{r}
# weights in kg, uniform distribution (100 units, a = 40, b = 120)
weights <- runif(n = 100, min = 40, max = 120)
# heights in cm, uniform distribution (100 units, a = 152, b = 214)
heights <- runif(n = 100, min = 152, max = 214)
```

### Put data into dataframe
```{r}
df <- data.frame(weights, heights)
```

\newpage

## Plot data
### weight (x-axis), height (y-axis)
```{r}
ggplot(
  data=df,
  aes(x=weights,
      y=heights
      )
) +
geom_point()
```

### height (x-axis), weight (y-axis)
```{r}
ggplot(
  data=df,
  aes(x=heights,
      y=weights
      )
) +
geom_point()
```

\newpage

## Model creation

### weights (X), heights (Y)
```{r}
model_a <- lm(heights ~ weights)
summary(model_a)
```

### heights (X), weights (Y)
```{r}
model_b <- lm(weights ~ heights)
summary(model_b)
```

\newpage

## Standardizing data
```{r}
z_heights <- scale(heights)
z_weights <- scale(weights)
```

## Model creation w/ standardized data
### z_weights (X), z_heights (Y)
```{r}
z_model_a <- lm(z_heights ~ z_weights)
summary(z_model_a)
```

### z_heights (X), z_weights (Y)
```{r}
z_model_b <- lm(z_weights ~ z_heights)
summary(z_model_b)
```
