---
title: "MCAT Modeling & Analysis"
author: "Tarek El-Hajjaoui"
date: "2023-02-25"
output: pdf_document
---

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Loading libraries
library(tidyverse)
library(janitor)
library(ggplot2)
```

### Data loading & pre-processing

Loading the dataset
```{r}
file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/R_Statistical_Modeling/MCAT/MedGPA.txt'
df = read.table(file_path, header=TRUE, sep="", dec=".")
```

Summary of dataset
```{r}
str(df)
```

Transforming categorical columns to factor data type.
```{r}
categorical_cols <- c('Accept', 'Acceptance', 'Sex')
df[categorical_cols] <- lapply(df[categorical_cols], as.factor)
```

Ensuring column data types are correct now.
```{r}
str(df)
```

\newpage

### Splitting up data to supplement analysis in 2-sample t-test

Separating the dataset into Female observations & Male observations
```{r}
df_female <- df[df$Sex == 'F',]
df_male <- df[df$Sex == 'M',]
```

Female dataset summary
```{r}
str(df_female)
```

Male dataset summary
```{r}
str(df_male)
```

### 2-sample t-test: Y=MCAT scores, split dataset on Sex (male or female)
```{r}
two_sample_t_test <- t.test(df$MCAT[df$Sex=="F"],df$MCAT[df$Sex=="M"], var.equal=TRUE)
two_sample_t_test
```

\newpage

#### Model 1
#### Y=MCAT scores, X1=GPA, X2=Sex (male or female)

```{r}
model <- lm(MCAT ~ GPA + Sex, data=df)
```

#### Summary of Model 1

```{r}
summary(model)
```

\newpage

#### Model 2 (adding interaction term, GPA*Sex)
#### Y=MCAT scores, X1=GPA, X2=Sex, X3=GPA*Sex

```{r}
model_2 <- lm(MCAT ~ GPA + Sex + GPA*Sex, data=df)
```

#### Summary of Model 2

```{r}
summary(model_2)
```

\newpage

#### Predictions

predicted MCAT score for a female with a 4.0 GPA:
```{r}
predict(model_2, data.frame(GPA=4.0, Sex='F'), se.fit=TRUE)
```

predicted MCAT for a male with a 4.0 GPA:
```{r}
predict(model_2, data.frame(GPA=4.0, Sex='M'), se.fit=TRUE)
```
