---
title: "Midwest Sales"
author: "Tarek El-Hajjaoui"
date: "2023-01-29"
output: pdf_document
---

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
# Loading libraries
library(tidyverse)
library(janitor)
```

Loading the dataset
```{r}
file_path = '/Users/Tarek/Documents/UCI_MDS_Coding/Stats210P/R_Statistical_Modeling/MidwestSales/MidwestSales.txt'
Midwest = read.table(file_path, header=FALSE, sep="", dec=".")
```

Give labels to the columns of the dataset.
```{r}
names(Midwest)=c("id","price","sqft","bed","bath","ac","garage","pool","year","quality","style","lot","hwy")
summary(Midwest)
```

\newpage

## Hypothesis Testing (two-sided test)
* Null Hypothesis: there is no significant linear relationship between the square footage and price of a house; (B1 = 0).
* Alternative Hypothesis: there is a significant linear relationship between the square footage and price of a house; (B1 != 0).

## Confidence Intervals

### Creating a Linear Model where the square footage of the house is used to predict the sale price.
```{r}
model <- lm(price ~ sqft, data=Midwest)
summary(model)
```

### Confidence Interval (CI) for sqft B0 value
```{r}
confint(model)
```

\newpage

### 95% confidence interval for the mean value of the response (price) when sqft=2000
```{r}
# interval="c" means confidence interval.
predict(model, list(sqft=2000), interval= "c")
```

### 95% prediction interval for the mean value of the response (price) when sqft=2000
```{r}
# interval="c" means confidence interval.
predict(model, list(sqft=2000), interval= "p")
```

### 90% prediction interval for the mean value of the response (price) when sqft=2000
```{r}
# interval="c" means confidence interval.
predict(model, list(sqft=2000), interval= "p", level = 0.90)
```
